---
title: "RMDL Assessment 2: Data Analysis Report"
authors:
  - name: B287980
bibliography: references.bib
execute: 
  echo: true
---

## **Emotional Stroop effects in French-English bilingual speakers**

## Background

The Stroop effect is a cognitive interference effect from the Stroop task [@stroop_studies_1935], whereby identification of a written colour word’s ink colour is slower when the written word is printed in a conflicting colour: for example, the word ‘blue’ printed in green ink is more slowly identified than when it is printed in blue link, due to interference from the automatic process of reading during colour identification. Subsequent research has led to several variations of the Stroop task. One such variant is the emotional Stroop task, in which participants identify the ink colour of emotionally salient and neutral words rather than colour words. The resulting interference effect is theorised to be the result of interference caused by the emotion stimuli, the process for which is automatic and immediate \[@bargh_automaticity_1997\]. In addition, the Stroop paradigm has been widely applied in bilingual contexts to examine patterns of interference across languages \[@magiste_stroop_1984\]. In bilinguals, research often reflects that the mother tongue (L1) is more strongly associated with emotion, while the second language (L2) can be used to create emotional distance [@dewaele_emotions_2010]. However, results also reflect that early simultaneous bilinguals demonstrate similar patterns of response to emotion words in both languages [@harris_when_2006], meaning processing of emotion words in L1 and L2 may vary with proficiency, with lower proficiency typically leading to a greater interference effect \[@sutton_automatic_2007\]. This may reflect reduced automaticity in lexical processing, leading to increased competition between linguistic and task-related processes. The following study builds on the theory of the emotional Stroop effect and bilingual Stroop effect, to investigate the effect of age of acquisition of emotional interference in English-French bilingual speakers. 

### Research Questions/Hypotheses

**Research question:** Does age of acquisition influence reaction time to emotion and to neutral words in a bilingual Stroop paradigm? \
\
**Research hypothesis:** Hypothesis is based on the results of @sutton_automatic_2007. In early bilinguals, a slower reaction time is predicted for emotion words compared to neutral words in both French and English conditions. For late bilinguals, a slower reaction time is predicted for emotion words than for neutral words only in the English condition. An interaction between word condition and language is therefore only expected in late bilinguals. Therefore, we expect that an emotion interference effect will negatively correlate with language proficiency, with early bilinguals showing faster reaction times than late bilinguals.

## Methods

### Participants

The study uses two participants groups: early sequential and simultaneous bilingual French-English speakers (attainment of both languages before the age of 11) and late sequential French-English speakers (native English speakers who learnt French as a second language after the age of 11). The late bilingual group had 6 participants, and the early bilingual groups had 5 participants, so a total of 11 participants took part in the study. 

In the late bilingual group, acquisition of French was between the ages of 11 - 15 (mean = 11), with fluency (mean = 71% range = 50 - 90%) achieved from the ages of 16 - 20 (mean = 18), with one participant describing themselves as not fluent in French. All participants had proficiency and/or fluency additionally in Italian, Spanish, German, and BSL or a combination of these languages. 1 participant had acquired Italian and English from birth, and had acquired L4 French after L3 German, while the remaining 5 participants had acquired French as an L2. On average, participants were exposed to English 79.5% of the time, French 8.4% of the time, and other languages 12.1% of the time. 4 participants reported that they would choose to read in English over another language, while 1 participant reported a 50% preference for French and English, and 1 participant reported that they would choose to read in Italian. 

In the early bilingual group, 4 participants were simultaneous learners of French and English, meaning they acquired them simultaneously from birth, while 1 participant acquired French from birth and English from age 4, and 1 participant was a native French speaker who acquired fluency in English from age 10. All participants had proficiency and/or fluency additionally in Spanish, German and Russian, or a combination of these languages. On average, participants were exposed to English 73.3% and to French 23% of the time, and to another language 3.7% of the time. 4 participants reported that they would choose to read in English, while 1 reported that they would choose to read in French. 

All participants were aged between 19 - 28 and are university educated, and currently reside in the United Kingdom or the United States.

#### Table 1 - Participant characteristics

+-------------------+--------------------------------------------+--------------------------------------------------+-------------------------------------------------------+---------------------------+
| Participant group | Age of acquisition of French (mean, range) | Self-rated fluency level in French (mean, range) | Age at which fluency in French achieved (mean, range) | Current language exposure |
+===================+============================================+==================================================+=======================================================+===========================+
| Late bilinguals   | 11, 11 - 15                                | 71%, 50 - 90%                                    | 18, 16 - 20                                           | English = 79.5%           |
|                   |                                            |                                                  |                                                       |                           |
|                   |                                            |                                                  |                                                       | French = 8.4%             |
|                   |                                            |                                                  |                                                       |                           |
|                   |                                            |                                                  |                                                       | Other = 12.1%             |
+-------------------+--------------------------------------------+--------------------------------------------------+-------------------------------------------------------+---------------------------+
| Early bilinguals  | 1, 0-4                                     | 100%                                             | N/A                                                   | English = 73.3%           |
|                   |                                            |                                                  |                                                       |                           |
|                   |                                            |                                                  |                                                       | French = 23%              |
|                   |                                            |                                                  |                                                       |                           |
|                   |                                            |                                                  |                                                       | Other = 3.7%              |
+-------------------+--------------------------------------------+--------------------------------------------------+-------------------------------------------------------+---------------------------+

### **Materials**

***Language history questionnaire***

Firstly, all participants completed a language history questionnaire [@marian_leap-questionnaire_2007], which collects information on their language acquisition background and self-proficiency ratings. The information is presented below in Figure 1 (above).

***Stroop task*** 

The Stroop task assessment was built using the materials adapted from @sutton_automatic_2007. In the original study, the word list was composed of two different types of words: negative emotion words (e.g. anger) and neutral (non-emotion related) words which belonged to the category ‘nautical nouns’ (e.g. boat). The English words were repeated in Spanish. @sutton_automatic_2007 highlight that the chosen words were high in arousal and low in valence, and matched for frequency and length. Therefore, the current study uses the same English word list but with French translated equivalents (rather than Spanish) in order to build on the results of @sutton_automatic_2007 and investigate a different language combination. The complete word list (French & English, emotion and neutral items) was repeated in blue ink and in green ink. The word list was therefore divided into three blocks: language (French/English), word type/condition (neutral/emotion) and ink colour (blue/green). Due to the small scale of the study, the translated French words were not matched for frequency and length. 

Participants saw a total of 32 items, as items were counterbalanced across colours and languages: half of the emotion items appeared in English and half in French, and the same applied to neutral items. The items were not repeated across languages (for example, if a participant saw the English word ‘boat’ they would not see the French equivalent ‘bateau’). Items were also counterbalanced across colour: each word was presented once, in one colour, and not repeated in the second colour. In summary, participants saw eight neutral English items (4 in green, 4 in blue), eight emotion English items (4 in green, 4 in blue), eight neutral French items (4 in green, 4 in blue), and eight emotion French items (4 in green, 4 in blue). Items were blocked by language, and the order that the blocks were presented in was pseudorandomised and counterbalanced across participants.

A complete list of stimuli can be found in the Appendix.

### **Procedure**

The Stroop task was self-conducted online via Testable, with each participant completing it on their own device. Participants first completed a practice session, consisting of a total of 6 neutral items, which were in a distinct semantic category to the critical trial neutral items. In the practice condition, the word appeared on the screen for 3 seconds. Participants were instructed to press the ‘L’ key if the word appeared in green ink, and the ‘A’ key if the word appeared in blue ink. Instructions only appeared in English. For the practice trials, the key/colour reminder prompt appeared on the screen alongside the practice word. Participants were not given feedback on the practice condition if their answer was correct. Following the practice condition, the participants immediately completed the critical trial conditions. In the critical condition, the stimuli appeared on the screen for less time - a total of 1.7 seconds - and the prompt did not appear on the screen alongside the stimuli.

### **Statistical Analysis**

In order to answer the research question, *does age of acquisition influence reaction time to emotion and to neutral words in a bilingual Stroop paradigm,* age of acquisition was treated as a categorical variable of early bilinguals and late bilinguals, and the bilingual nature of the Stroop paradigm is operationalised as language of stimuli (English, French).

Reaction times in milliseconds for each trial will be analysed using mixed-effects regression modelling. Summary measures for mean and standard deviation of the raw data will be reported. The study will then employ a 2x2x2 mixed factorial design of lognormal regression, with group (early vs late) as between-participant fixed effects and condition (emotion vs neutral) and language of stimuli (French vs English) as within-participant random slopes. The interactions between groups and conditions will also be analysed.

### **Data Availability**

Anonymised participant data (Stroop task performance) is available on the following GitHub link: <https://github.com/KatiePedrotti/rmdl-template>

To respect participant privacy, language questionnaires are not available.

## Results: Data analysis

### Reading and processing the data

```{r}
#| label: read-data
library(tidyverse)
library(ggplot2)
library(brms)
library(dplyr)
library(posterior)
library(tidybayes)

```

```{r}
alldata = list.files(
  path = "./data/",
  pattern = "*.csv$",
  recursive = TRUE ,
  full.names = TRUE
  
)

compiledlist <- lapply(alldata, function(x) {
  read_csv(x, skip = 3)
})

finaldata <- bind_rows(compiledlist) |>
  filter(condition1 != "practise") |>
  mutate(
    early_late = case_when(
      substring(id,1,1) == "1"~"early",
      substring(id,1,1) == "2"~"late",
      TRUE ~""
    )
  )
```

As the data were in separate files and lacking identifying columns, I firstly used list.files() to list all the available results files, and lapply () to compile them together, with read_csv(x, skip =3) in order to skip the first 3 rows of each dataset which did not contain any data. Then I used bind_rows to read in this data, and filtered out the 6 practise conditions using filter(). I then used mutate() in order to create a column which classified the participants as early or late bilinguals based on their participant number (1 were early bilinguals, 2 were late bilinguals).

```{r}
newfinaldata <- finaldata |>
  separate_wider_delim(
    condition1,
    delim = "_",
    names = c("Condition", "Language")
  )
```

```{r}
newfinaldata$Language <- recode(newfinaldata$Language,
                             "eng" = "English",
                             "fr"  = "French")

newfinaldata$early_late <- recode(newfinaldata$early_late,
                               "early" = "Early bilingual",
                               "late"  = "Late bilingual")

newfinaldata$Condition <- recode(newfinaldata$Condition,
                               "emotion" = "Emotion",
                               "neutral"  = "Neutral")
```

I used separate_wider_delim() in order to create new columns which identified the condition (emotion and neutral) and language (French and English), and re-coded the column names accordingly. The relevant columns are therefore Condition (emotion/neutral stimuli), Language (French/English stimuli), and Early_Late (early or late bilingual), as well as RT (reaction time).

### Plotting the raw data

```{r}
newfinaldata |>
  ggplot(aes(early_late, RT, colour = Language)) +
  geom_jitter(alpha = 0.7, width = 0.1) +
  facet_wrap(~ Condition) +
  labs(
    x = "Early or Late Bilingual", 
    y = "Reaction Time (ms)", 
    title = "AoA (Early or Late Bilingual) influence on reaction time 
    to emotion and to neutral words")
```

#### Figure 1 - Age of acquisition (early or late bilingual) effect on reaction time (RT) across condition (emotion and neutral stimuli) and language (French and English). 

I used a jitter plot to represent the raw data, since it combines numeric (reaction times) and categorical variables. The jitter plot is faceted by Condition (neutral and emotion stimuli), with the x-axis showing Participant group (early or late bilingual) and the y-axis showing Reaction Time (ms). The colours differentiate the Language of the stimuli (French = blue, English = red). 

The plot shows firstly in the emotion condition, that late bilinguals showed less variation and therefore are more densely represented than the early bilinguals. There is minimal variation across both participant groups between languages (English compared to French). 

In the neutral condition, there is slightly more variation across languages, with late bilinguals appearing to perform very slightly faster in English than in French, and early bilinguals' reaction times showing more variation compared to late bilinguals in this condition. The plot reflects a slightly faster reaction time across conditions and across language for late bilinguals.

There are some outliers across condition, group and language, but these appear to be more pronounced in the early bilinguals comapred to the late bilinguals.

The raw data is further described in the summary measures (below).

### Summary of the raw data

```{r}
newfinaldata |> 
  group_by(Condition, Language, early_late) |> 
  summarise(
    mean = mean(RT), sd = sd(RT))
```

In order to investigate if age of acquisition affects reaction time to emotion and neutral words in the bilingual Stroop task, I analysed participant reaction times across condition (neutral vs. emotion) and language (French vs. English). 

Across conditions, late bilinguals showed faster mean reaction times than early bilinguals. 

In the emotion condition, early bilinguals had a faster reaction time in French (mean = 726.38ms, SD = 440.26) than in English (mean = 817.83ms, SD = 561.09), while the late bilinguals showed the reverse pattern, demonstrating slightly faster reaction times in French (mean = 533.33ms, SD = 215.51) than in English (mean = 542.31ms, SD = 255.28). Overall in the emotion condition, the late bilinguals had faster reaction times in both languages than the early bilinguals. 

A similar pattern can be observed in the neutral condition. In language conditions, early bilinguals had a faster reaction time in French (mean = 755.85ms, SD = 376.74ms) than in English (mean = 805.58ms, SD = 536.91ms) in the neutral condition, while late bilinguals had a faster reaction time in English (mean = 515.77ms, SD = 196.49ms) than in French (mean = 536.77ms, SD = 204.00ms) in the neutral condition. Therefore overall in the neutral condition, late bilinguals showed faster reaction times in both languages compared to early bilinguals. 

Comparing conditions, early bilinguals showed a slightly faster reaction time in the emotion condition in English (English: mean = 817.83ms, SD = 561.08ms) compared to the neutral condition in English (English: mean = 805.58ms, SD = 536.91ms) but showed the reverse pattern in the emotion condition in French (emotion: mean = 726.38ms, SD = 440.26, neutral: mean = 755.85ms, SD = 376.74ms). 

The late bilinguals showed the reverse pattern, demonstrating a slower reaction time in the emotion condition in English (mean = 542.31, SD = 255.28) compared to the neutral condition in English (mean = 515.77ms, SD = 196.49ms), but the same pattern as the early bilinguals across the French conditions, such that they performed slightly faster in the French emotion condition (mean = 533.33ms, SD = 215.51) then the French neutral condition (mean = 536.77ms, SD = 204.00ms), but to a lesser extent than the early bilinguals demonstrated. 

Taken together, late bilinguals had faster reaction times than early bilinguals, with early bilinguals showing a faster reaction time to emotion words in French (compared to neutral condition), and late bilinguals showing a faster reaction time in the neutral condition in English. The differing patterns across participant groups in the raw data therefore suggest that age of acquisition may affect reaction time to emotion and neutral words across condition and language, such that late bilinguals showed a faster reaction time across both language and condition.

### Regression modelling

I used Bayesian log-normal regression model for the outcome variable reaction time (RT) using brms. This distribution allows positive-only values, continuous variables, and is able to account for right-skewed data. This was fitted with Group (early vs late bilinguals) as between-participant fixed effects, and Condition (emotion vs neutral) and Language (French vs English) as within-participant random effects, and including Participant ID as random effect to account for inter-participant variability. I then obtained summary measures. As the values are reported as logged reaction time, I used exp() on the relevant values to obtain the unlogged reaction times.

```{r}
brm_newfinaldata <- brm(
  RT ~ early_late * Language * Condition +
    (Language + Condition | id ),
  family = lognormal,
  data = newfinaldata,
  seed = 456,
  cores = 4,
  file = "cache/brmfinaldata", 
  control = list(adapt_delta = 0.99, max_treedepth = 15)
)
```

```{r}
summary(brm_newfinaldata)
```

```{r}
exp(6.17)  #Intercept_95%
exp(6.92)  #Intercept_95%
```

The model did converge, however the summary measures reflect a high degree of uncertainty. 

At 95% credible internal, the model estimates that the mean reaction time for early bilinguals in the emotion and English condition is between 6.17 (lower bound) and 6.92 (upper bound) logged milliseconds, and 478.19ms (lower bound) and 1,0312.32ms (upper bound) unlogged milliseconds. All predictors cross zero at 95% credible internal, and therefore there is a high degree of uncertainty as to their effect size. 

```{r}
summary(brm_newfinaldata, prob = 0.8)
```

```{r}
exp(6.31)  #Intercept_80%
exp(6.76)  #Intercept_80%
```

At 80% credible interval, all predictors still cross zero, but to a lesser extent. The mean reaction time predicted at the Intercept (english_emotion_early) is between 6.31 logged milliseconds and 6.76 logged milliseconds, and 550.04 and 862.64 unlogged milliseconds.  

```{r}
summary(brm_newfinaldata, prob = 0.6)
```

```{r}
exp(6.39)  #Intercept_60%
exp(6.67)  #Intercept_60%
exp(-0.49) #early_lateLatebilingual_60%
exp(-0.09) #early_lateLatebilingual_60%
exp(-0.14) #LanguageFrench
exp(-0.03) #LanguageFrench
```

Finally, at 60% credibility, the predicted reaction time at the intercept is between 6.17 and 6.92 logged milliseconds, or between 595.86 and 788.38 unlogged milliseconds (english_emotion_early). At 60% credibility, two predictors now show a decrease in log milliseconds. Firstly, english_emotion_late show an effect size between -0.49 and -0.09 (logged) or 0.61 to 0.91 (unlogged), suggesting that late bilinguals are predicted to be faster than early bilinguals in the emotion/English condition. Secondly, french_emotion_early show an effect size between -0.14 and -0.03 (logged) or 0.86 to 0.97 (unlogged), suggesting that early bilinguals in the emotion condition are likely to show a faster reaction time in French than in English. However, the remaining predictors cross zero and there remains a high degree of uncertainty regarding how these predictors interact across the model. 

After obtaining summary measures at 95%, 80% and 60% credible intervals, I obtained the model draws, implementing the relevant predictors and interactions based on the b_Intercept as english_emotion_early. I then had select() the relevant columns and pivot_longer() in order to make new columns for these predictors, which I could then plot into the graph (below).

```{r}
draws_brm_newfinaldata <- as_draws_df(brm_newfinaldata)

frame_draws_brm_newfinaldata <- draws_brm_newfinaldata |>
  mutate(
    eng_emotion_early = b_Intercept, 
    fr_emotion_early = b_Intercept + b_LanguageFrench,
    eng_emotion_late = b_Intercept + b_early_lateLatebilingual, 
    eng_neutral_early = b_Intercept + b_ConditionNeutral, 
    fr_emotion_late = b_Intercept + b_early_lateLatebilingual
    + b_LanguageFrench + `b_early_lateLatebilingual:LanguageFrench`,
    eng_neutral_late = b_Intercept + b_ConditionNeutral + 
      b_early_lateLatebilingual + 
      `b_early_lateLatebilingual:ConditionNeutral`,
    fr_neutral_early = b_Intercept + b_LanguageFrench + 
      b_ConditionNeutral +
      `b_LanguageFrench:ConditionNeutral`,
    fr_neutral_late = b_Intercept + b_ConditionNeutral + 
      b_LanguageFrench +
      b_early_lateLatebilingual + 
      `b_early_lateLatebilingual:LanguageFrench` + 
      `b_LanguageFrench:ConditionNeutral` + 
      `b_early_lateLatebilingual:ConditionNeutral` +
      `b_early_lateLatebilingual:LanguageFrench:ConditionNeutral`
  ) |>
  select(eng_emotion_early, fr_emotion_early, eng_emotion_late, 
         eng_neutral_early, fr_emotion_late, eng_neutral_late, 
         fr_neutral_early, fr_neutral_late)

```

```{r}
plot2 <- frame_draws_brm_newfinaldata |>
  select(eng_emotion_early, fr_emotion_early, eng_emotion_late, 
         eng_neutral_early, fr_emotion_late, eng_neutral_late, 
         fr_neutral_early, fr_neutral_late) |>
pivot_longer(
  eng_emotion_early:fr_neutral_late, 
  names_to = "cond", 
  values_to = "epred"
) |>
separate(cond, c("Language", "Condition", "Early_Late"))
  
  
```

```{r}
plot2 |>
  ggplot(aes(epred, Condition, fill = Language )) +
  stat_halfeye(alpha = 0.7) +
  scale_fill_manual(
    labels = c("English", "French"),
    values = c("lightskyblue2", "slateblue1") 
  ) + facet_grid(cols = vars(Early_Late)) +
  labs(title = "Reaction times (logged) across conditions 
       for early and late bilinguals", x = "Reaction times (logged ms)")
  
  
```

#### Figure 2 - Age of acquisition (early or late bilingual) effect on logged reaction time (RT) across condition (emotion and neutral stimuli) and language (French and English). 

I used a half-eye plot to demonstrate logged Reaction times on the x-axis, faceted by Participant group (early, late), and Condition on the y-axis (neutral, emotion), with colour differentiating language (English, French). The plot reflects the summary measures and model draws, showing very little variability between performance of early and late bilinguals across condition and across language. The most noticeable variation is in the emotion condition for early bilinguals, who appear to show a slightly faster reaction time in French than in English: the model did predict a negative effect on reaction time (i.e. faster reaction time) in these conditions at 60% credibility, however the summary measures showed this to still have a large degree of uncertainty.

## Discussion

**Research question:** Does age of acquisition influence reaction time to emotion and to neutral words in a bilingual Stroop paradigm? 

The research question asked does age of acquisition influence reaction times across conditions (emotion vs neutral) and across language (French vs English). In other words, is an emotional Stroop interference effect observed across different age of acquisition groups (early vs late bilinguals) and when the stimuli is presented in an L1 or an L2. 

While analysis of the raw data suggested that late bilinguals show a faster reaction time than early bilinguals across conditions (emotion vs neutral) and across languages (French vs English), the statistical analysis returned highly uncertain results and therefore no conclusions can be drawn at this stage.

### Limitations

It is worth mentioning several limitations of the study. Firstly, it is still a small participant cohort (5 early bilinguals and 6 late bilinguals), which may have contributed to the high  degree of certainty in the results. Secondly, operationalising age of acquisition as a categorical variable may have obscured between-group differences ( [@singleton_age_2022; @wagner_what_2022], and therefore treating age of acquisition as a continuous variable may have led to different results. Additionally, since the main focus of the study was investigating the emotional Stroop effect, rather than the Stroop effect associated with orthographical/colour interference, I chose to still include the incorrect trials alongside the correct trials, which may have contributed to the variation in results. Finally, it is likely that the stimuli presented some problems, since the stimuli used by Sutton (2007) and replicated in this study contain obscure neutral nouns of boat parts, which are uncommon both in English and in French. It is possible that participants (in particular the late bilingual group, who had lower levels of proficiency in French than the early bilingual group) were unfamiliar with the neutral lexical items, which may have conversely affected their reaction times compared to a different set of more common neutral lexical items. This may therefore suggest that results could be due to proficiency levels, rather than age of acquisition.

## Appendix

### Stimuli 

| English - Emotion | French - Emotion | English - Neutral | French - Neutral  |
|-------------------|------------------|-------------------|-------------------|
| nervous           | nerveux          | engine            | barre             |
| grief             | chagrin          | boom              | bôme              |
| upset             | contrarié        | stern             | poupe             |
| greed             | avidité          | cabin             | cabine            |
| anxiety           | anxiété          | keel              | quille            |
| jealousy          | jalousie         | motor             | moteur            |
| lonely            | seul             | seat              | siège             |
| scared            | effrayé          | deck              | pont              |
| afraid            | craintif         | driver            | conducteur        |
| panic             | panique          | propoeller        | hélice            |
| fear              | peur             | floor             | plancher          |
| angry             | énervé           | anchor            | ancre             |
| disgusted         | dégoûté          | paddle            | pagaie            |
| depressed         | déprimé          | beam              | barrot            |
| mad               | furieux          | port              | bâbord            |
| troubled          | tourmenté        | rope              | corde             |
